---
title: "Downstream fill taxa traits"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(oharac) ### remotes::install_github('oharac/oharac')
library(data.table)
library(tidyverse)
library(here)
library(readxl)
source(here('common_fxns.R'))
library(taxize)

```

# Summary

Using taxonomic trees from WoRMS and the `taxize` package, we will fill traits for downstream non-scored species (i.e., species whose partial traits were extracted from FishBase/SeaLifeBase and range maps).

# Methods

* Load species traits provided by experts and filled with additional information from FishBase/SeaLifeBase   
* Join the known species traits to the expanded taxonomic list to get the full classification.
* For traits known at a non-species level, we will downfill those to the species level.

## Assemble the taxonomic dataframe

Grab data from each rank (kingdom --> phylum, phylum --> class, class --> order, etc.) and assemble into a long dataframe.  Long format will be easier to match trait-scored species to, since all ranks will be in the "name" column.  But wide format will be easier for keeping track of the structure of each species' classification.

Because we are dropping plants and algae, filter wide dataframe to just species in kingdom _Animalia_.  This is done in the `assemble_worms()` function from `common_fxns.R`.

``` {r assemble full spp taxonomic table from WoRMS}
spp_all_wide <- assemble_worms('wide', seabirds_only = TRUE)
spp_all_long <- assemble_worms('long', seabirds_only = TRUE)
```

## Match species with known traits to most specific rank

Load the trait data, and match to the wide-format species classifications along with the representativeness corrections, to identify species that represent higher ranks.  Create a separate dataframe of these higher ranks, based on relative frequency of traits across species within that higher rank.

First step, expand the spp group traits to complete the trait set for all species, especially those with only partial traits from FB and range filling.  

```{r expand traits across spp}
spp_gp_traits <- read_csv(here('_data/spp_traits_filled.csv'))

spp_to_taxa <- spp_gp_traits %>%
  select(spp_gp, taxon) %>%
  distinct()

spp_gp_traits_complete <- spp_gp_traits %>%
  select(-taxon) %>%
  complete(spp_gp, nesting(category, trait)) %>%
  left_join(spp_to_taxa, by = 'spp_gp') %>%
  mutate(note = 'traits expanded')
```


``` {r determine representative spp gps}

repres_fish <- read_excel(here('_raw_data/xlsx/fish_traits_add_bp.xlsx'), 
                          sheet = 'repres_expanded')
repres_df   <- read_excel(here('_raw_data/xlsx/spp_gp_representativeness.xlsx')) %>%
  janitor::clean_names() %>%
  select(taxon, species, repres = representative_rank) %>%
  bind_rows(repres_fish) %>%
  mutate(species = tolower(species),
         repres = tolower(repres)) %>%
  filter(!is.na(repres) & repres != 'species')

### create new dataframe including all representative species, with
### spp_gp_new assigned to the representative rank
repr_spp <- spp_gp_traits %>%
  inner_join(repres_df, by = c('taxon', 'spp_gp' = 'species')) %>%
  inner_join(spp_all_wide, by = c('spp_gp' = 'species')) %>%
  mutate(spp_gp_new = case_when(repres == 'genus'  ~ genus,
                                repres == 'family' ~ family,
                                repres == 'order'  ~ order,
                                repres == 'class'  ~ class,
                                TRUE ~ NA_character_))

non_mut_excl <- c('adult_mobility', 'biomineral', 'zone',
                  'calcium_carbonate_structure_location', 
                  'depth_min_max', 'respiration_structures', 
                  'across_stage_dependent_habitats_condition', 
                  'within_stage_dependent_habitats_condition')

### group by the new spp gp and summarize traits
repr_spp_gps <- repr_spp %>%
  group_by(taxon, spp_gp_new, category, trait) %>%
  mutate(tot_prob = sum(trait_prob),
         tot_spp = n_distinct(spp_gp)) %>%
  group_by(taxon, spp_gp_new, category, trait, trait_value) %>%
  summarize(trait_prob = case_when(
              first(trait) %in% non_mut_excl ~ n() / first(tot_spp),
              TRUE ~ sum(trait_prob) / first(tot_prob)),
            nspp = n_distinct(spp_gp)) %>%
  ungroup() %>%
  rename(spp_gp = spp_gp_new) %>%
  mutate(note = 'repr groups')

### to check functioning of non-mutually-exclusive trait calcs
# x <- repr_spp_gps %>% filter(trait %in% non_mut_excl)
# y <- x %>% group_by(spp_gp, category, trait) %>% summarize(tot_prob = sum(trait_prob))
```

Then bind the representative-based dataframe with the original traits data, and join to long-format species classifications.

``` {r match up}
### add representative spp gps to the original spp_gp_traits. Note
### this will keep the species level scores intact, while also adding
### higher order representative ranks.

spp_gp_traits_repr_fix <- spp_gp_traits_complete %>%
  bind_rows(repr_spp_gps) %>%
  select(-nspp, -n_spp_gps) %>%
  clean_scinames('spp_gp')

matched_df <- spp_all_long %>%
  left_join(spp_gp_traits_repr_fix, by = c('name' = 'spp_gp')) %>%
  mutate(match = ifelse(is.na(trait_value), NA, as.character(rank))) %>%
  filter(!is.na(trait_value)) %>%
  group_by(spp, trait) %>%
  filter(as.numeric(rank) == max(as.numeric(rank))) %>%
  ungroup()
### matched_df$rank %>% unique()
### [1] class   order   family  genus   species
### Levels: kingdom phylum class order family genus species

```

## Check species counts

Check species counts by taxon, excluding spp with only 4 or fewer traits

```{r check spp counts}
check_spp <- matched_df %>%
  group_by(spp) %>%
  filter(n_distinct(trait) > 4) %>%
  mutate(n_trait = n_distinct(trait)) %>%
  group_by(taxon) %>%
  summarize(mean_n_traits = round(mean(n_trait), 2),
            n_spp = n_distinct(spp)) %>%
  arrange(desc(n_spp))

knitr::kable(check_spp)

### sum(check_spp$n_spp)
### 30030 spp - same as before - since anything with more than four traits
### clearly already had traits in the excel sheets...
```

## Write out results

Break into several chunks to reduce duplication of info and resulting file size...

```{r write out results}
match_taxa <- matched_df %>%
  select(taxon, spp) %>%
  distinct() %>%
  arrange(taxon, spp) %>%
  mutate(df_spp_id = 1:n())

match_traits <- matched_df %>%
  select(category, trait, trait_value) %>%
  distinct() %>%
  arrange(category, trait, trait_value) %>%
  mutate(df_trait_id = 1:n())

match_levels <- matched_df %>%
  select(match, match_name = name) %>%
  distinct() %>%
  arrange(match, match_name) %>%
  mutate(df_match_id = 1:n())

match_spp_traits <- matched_df %>%
  left_join(match_taxa, by = c('taxon', 'spp')) %>%
  left_join(match_traits, by = c('category', 'trait', 'trait_value')) %>%
  left_join(match_levels, by = c('match', 'name' = 'match_name')) %>%
  select(df_spp_id, df_trait_id, df_match_id, trait_prob) %>%
  distinct()
  
write_csv(match_taxa,       
          here_anx('1_downfill_traits/spp_match_downfill_taxa.csv'))
write_csv(match_traits,     
          here_anx('1_downfill_traits/spp_match_downfill_traits.csv'))
write_csv(match_levels,     
          here_anx('1_downfill_traits/spp_match_downfill_levels.csv'))
write_csv(match_spp_traits, 
          here_anx('1_downfill_traits/spp_match_downfill_spp_traits.csv'))

### to reassemble:
# match_df <- fread(here_anx('1_downfill_traits/spp_match_downfill_spp_traits.csv')) %>%
#   full_join(fread(here_anx('1_downfill_traits/spp_match_downfill_levels.csv'))) %>%
#   full_join(fread(here_anx('1_downfill_traits/spp_match_downfill_traits.csv'))) %>%
#   full_join(fread(here_anx('1_downfill_traits/spp_match_downfill_taxa.csv'))) %>%
#   select(-starts_with('df_match'))
              
```

