---
title: 'Align species names and AphiaID'
author: "Casey O'Hara"
format:
  html:
    embed-resources: true
---

## Summary

The spp vulnerability project used scientific binomials as the identifier.  WoRMS uses an AphiaID numeric identifier for different taxonomic ranks, and AquaX has taken on that identifier for mapping SDMs.

For this project, we should identify the AphiaID value for each of the species found within the spp vulnerability dataset.

* Match species names directly, comparing the vulnerability dataset to the AquaX META dataset.
* Identify species names that do not have a direct match; query in WoRMS (using `taxize` or similar).
* Save a disambiguated set of species names to AphiaID for use throughout the project.

## Data

Species vulnerability scores and trait values are presented in Butt et al. (2022). A trait-based framework for assessing the vulnerability of marine species to human impacts. Ecosphere, 13(2), e3919. https://doi.org/10.1002/ecs2.3919.

* Note, the species trait/vulnerability scores were recalculated after publication; code is available at https://github.com/mapping-marine-spp-vuln/spp_vuln_framework
* Updated traits and vulnerability scores will be available in KNB eventually.

AquaX species distribution models are presented in Reygondeau et al. (2026). AquaX: An enhanced and revised AquaMaps framework to model marine species distributions and biodiversity. PLOS ONE, 21(2), e0335823. https://doi.org/10.1371/journal.pone.0335823

* Data for 12k IUCN-listed species were provided from Gabriel Reygondeau to Casey O'Hara in April 2026, and placed on Aurora.

## Methods

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(taxize) ### remotes::install_github('ropensci/taxize')
source('common_fxns.R')
```

### Read in data

Read in data for species from vulnerability project and AquaX species.

```{r}
spp_traits <- data.table::fread(here_anx('2_gapfill_traits/spp_up_down_gapfill_taxa.csv')) %>%
  select(-starts_with('gf_')) %>%
  rename(sciname = species)
```

```{r}
load(here('_data/META_02122025.Rdata'))
spp_aquax <- META %>%
  janitor::clean_names() %>%
  select(aphia_id, sciname = scientific_name, kingdom:genus, sdm) %>%
  mutate(across(where(is.character), tolower)) %>%
  distinct()
```

Note, here we are using the full AquaX metadata, which includes ~170k spp with no SDM in the dataset.  However, these species are listed with their AphiaID so we can grab those values easily even if no map is available.

### Identify direct matches

This subset includes any species whose sciname is included in AquaX and thus tied to an AphiaID.

```{r}
spp_direct_match <- inner_join(spp_traits, spp_aquax, by = 'sciname')

nrow(spp_direct_match)
```

### Identify non-matches

```{r}
spp_nomatch <- anti_join(spp_traits, spp_aquax, by = 'sciname')
```

Of the ~7k species listed in the traits data that do not have a match in the AquaX data, query WoRMS to identify AphiaID.  Why are these not matched?

### Pass 1: Resolve non-matches

Try all names against WoRMS to match IDs directly using `get_wormsid()`.  This queries each name one at a time, so takes a while.  Compare results against direct matches above - `get_wormsid()` uses fuzzy matching so there may be some overlap with spp noted above, or spp with matches in AquaX already.

```{r}
pass1_f <- here('tmp/spp_aphia_id_pass1.csv')

if(!file.exists(pass1_f)) {
  check_classification <- function(x) {
    y <- classification(x, db = 'worms', ask = FALSE)
    if(is.data.frame(y[[1]])) {
      z <- y[[1]] %>%
        filter(rank == 'Species') %>%
        select(sciname_match = name, aphia_id = id) %>%
        mutate(sciname_query = x)
      return(z)
    } else {
      return(data.frame(sciname_query = x, aphia_id = NA))
    }
  }

  spp_vec <- spp_nomatch$sciname
  
  pass1_list <- lapply(X = spp_vec,
                       FUN = check_classification)
  
  pass1_df <- pass1_list %>%
    bind_rows()
  
  write_csv(pass1_df, pass1_f)
}

pass1_df <- read_csv(pass1_f)
```

Examine matches, including IDs that are already included in AquaX that didn't have a direct match from the start.

```{r}
sum(!is.na(pass1_df$aphia_id))

check1 <- inner_join(spp_aquax, pass1_df, by = c('aphia_id'))

DT::datatable(check1)
```

It appears that for the small overlapping subset, AquaX has the correct scientific names for the various Aphia ID values that are close matches.

### Pass 2?

```{r}
still_no_match <- pass1_df %>% filter(is.na(aphia_id))
```

Some of these are no longer extant! But `taxize` has no parameter to allow `extant = false`.

```{r}
### use %% to escape a true %
url_stem <- 'https://marinespecies.org/rest/AphiaRecordsByNames?scientificnames%%5B%%5D=%s&like=false&marine_only=true&extant_only=false'

pass2_f <- here('tmp/spp_aphia_id_pass2.csv')

if(!file.exists(pass2_f)) {
  
  check_extant <- function(x) {
    ### x <- unmatched_vec[1]
    spp_url <- sprintf(url_stem, x)
    y <- try(jsonlite::fromJSON(spp_url), silent = TRUE)
    if(is.data.frame(y[[1]])) {
      z <- y[[1]] %>%
        janitor::clean_names() %>%
        rename(sciname_match = scientificname) %>%
        mutate(sciname_query = x)
      return(z)
    } 
    ### if try-error?
    if(class(y) == 'try-error') {
      return(data.frame(sciname_query = x, aphia_id = NA, try_error = TRUE))
    }
    ### no error just empty?
    return(data.frame(sciname_query = x, aphia_id = NA, try_error = FALSE))
    
  }

  unmatched_vec <- still_no_match$sciname_query %>% str_replace(' +', '%20')

  pass2_list <- lapply(X = unmatched_vec,
                       FUN = check_extant)
  
  pass2_df <- pass2_list %>%
    bind_rows()
  
  write_csv(pass2_df, pass2_f)
}

pass2_df <- read_csv(pass2_f) %>%
  select(aphia_id, sciname_query, sciname_match, 
         valid_aphia_id, valid_name, 
         status, unacceptreason, is_extinct, try_error)
```

```{r}
### Still not found:
sum(pass2_df$try_error, na.rm = TRUE)

### Extinct:
sum(pass2_df$is_extinct, na.rm = TRUE)

### Accepted:
DT::datatable(pass2_df %>% filter(status == 'accepted' & !is_extinct))

### Not accepted:
DT::datatable(pass2_df %>% 
  filter(!is_extinct) %>%
  filter(status != 'accepted')
)

```

### Combined species/ID updates

```{r}
aquax_matched <- spp_direct_match %>%
  select(sciname, aphia_id) %>%
  mutate(valid_sciname = tolower(sciname))
pass1_matched <- pass1_df %>%
  filter(!is.na(aphia_id)) %>%
  mutate(sciname = sciname_query,
         valid_sciname = tolower(sciname_match)) %>%
  mutate(match = sciname == valid_sciname)

pass2_matched <- pass2_df %>%
  mutate(sciname_query = sciname_query %>% str_replace('.20', ' ')) %>%
  filter(!is.na(aphia_id)) %>%
  ### if no flag for extinct, keep it for now
  filter(is_extinct != 1 | is.na(is_extinct)) %>%
  mutate(valid_sciname = tolower(valid_name),
         aphia_id = valid_aphia_id) %>%
  mutate(match = sciname_query == valid_sciname) %>%
  select(sciname = sciname_query, valid_sciname, aphia_id, match) %>%
  distinct()

pass2_extinct <- pass2_df %>%
  filter(is_extinct == 1)
  
update_sciname_aphia_df <- bind_rows(aquax_matched, pass1_matched, pass2_matched) %>%
  select(sciname, aphia_id, valid_sciname) %>%
  distinct()
write_csv(update_sciname_aphia_df, 'update/update_sciname_aphia.csv')
```


## Reform outputs

Create updated outputs with simplified column names and file names; zip; put in .gitignore.

These will be uploaded to KNB when it becomes available.

### Trait outputs

Load dataframes of updated traits-to-species.  As aphia_id and valid_sciname may join two spp previously considered separate, and divide one spp previously considered unique, the trait_vals may have to change as well.

Recall that some traits are mutually exclusive (e.g., body length probs must add to 1) and others are not (e.g., habitats/zones may add up to more than one).

```{r}
library(data.table)
junction_df     <- fread(here_anx('2_gapfill_traits/spp_up_down_gapfill_spp_traits.csv')) %>%
  rename(trait_id = gf_trait_id, spp_id = gf_spp_id)
taxa_lookup     <- fread(here_anx('2_gapfill_traits/spp_up_down_gapfill_taxa.csv')) %>%
  rename(spp_id = gf_spp_id)
traitval_lookup <- fread(here_anx('2_gapfill_traits/spp_up_down_gapfill_traits.csv')) %>%
  rename(trait_id = gf_trait_id)
### these need reordering for sensible levels
gapfill_lookup  <- fread(here_anx('2_gapfill_traits/spp_up_down_gapfill_levels.csv')) %>%
  mutate(match = factor(match, levels = c('species', 'genus', 'family', 'order', 'class')),
         gapfill = factor(gapfill, levels = c('none', 'genus', 'family', 'order', 'class'))) %>%
  arrange(match, gapfill) %>%
  mutate(gf_match_id_new = 1:n())

junct_taxa_df <- full_join(junction_df, taxa_lookup, by = 'spp_id') %>%
  left_join(update_sciname_aphia_df, by = c('species' = 'sciname')) %>%
  left_join(traitval_lookup, by = 'trait_id') %>%
  left_join(gapfill_lookup, by = 'gf_match_id') %>%
  select(-spp_id, -species) %>%
  filter(!is.na(aphia_id))

### resolve trait inconsistencies due to many-to-many relationships among aphia_id and valid_sciname
non_mut_excl <- c('adult_mobility', 'biomineral', 'zone',
                  'calcium_carbonate_structure_location', 
                  'depth_min_max', 'respiration_structures', 
                  'across_stage_dependent_habitats_condition', 
                  'within_stage_dependent_habitats_condition')

resolve_mut_excl <- junct_taxa_df %>%
  filter(!trait %in% non_mut_excl) %>%
  group_by(aphia_id, category, trait) %>%
  ### keep only the most specific match level
  filter(gf_match_id_new == min(gf_match_id_new)) %>%
  ### total the probabilities across all observed values within the trait...
  mutate(sum_probs = sum(trait_prob), gf_ids = n_distinct(gf_match_id)) %>%
  ### then distribute trait values proportionally
  group_by(aphia_id, category, trait, trait_value, trait_id, gf_match_id = gf_match_id_new, match, gapfill) %>%
  summarize(trait_prob = sum(trait_prob) / sum_probs, .groups = 'drop')
  
resolve_nonmut_excl <- junct_taxa_df %>%
  filter(trait %in% non_mut_excl) %>%
  group_by(aphia_id, category, trait) %>%
  ### keep only the most specific match level
  filter(gf_match_id_new == min(gf_match_id_new)) %>%
  distinct() %>%
  ### no multiple instances with unique probability values
  select(aphia_id, category, trait, trait_value, trait_id, gf_match_id = gf_match_id_new, match, gapfill)

resolved_traits <- bind_rows(resolve_mut_excl, resolve_nonmut_excl)

### reassembled original dataset for verification:
traits_check_df <- fread(here_anx('2_gapfill_traits/spp_up_down_gapfill_spp_traits.csv')) %>%
  full_join(fread(here_anx('2_gapfill_traits/spp_up_down_gapfill_levels.csv'))) %>%
  full_join(fread(here_anx('2_gapfill_traits/spp_up_down_gapfill_traits.csv'))) %>%
  full_join(fread(here_anx('2_gapfill_traits/spp_up_down_gapfill_taxa.csv'))) %>%
  select(-starts_with('gf_')) %>%
  filter(species %in% junct_taxa_df$valid_sciname)
```

The new data include a few extra species (once unrecognized spp are dropped).

Break the resulting traits dataframe back into components to save out with new IDs.

```{r}
new_junct_df <- resolved_traits %>%
  select(aphia_id, trait_id, trait_prob, gf_match_id) %>%
  distinct()
new_trait_df <- resolved_traits %>%
  select(trait_id, category, trait, trait_value) %>%
  distinct() %>% arrange(trait_id)
new_taxa_df <- junct_taxa_df %>%
  select(aphia_id, sciname = valid_sciname, taxon) %>%
  distinct() %>% arrange(aphia_id)
new_gapfill_df <- resolved_traits %>%
  select(gf_match_id, gapfill, match) %>%
  distinct() %>% arrange(gf_match_id)

dir.create(here('update/spp_vuln_framework_traits'))
write_csv(new_junct_df, here('update/spp_vuln_framework_traits/spp_traits_junction.csv'))
write_csv(new_trait_df, here('update/spp_vuln_framework_traits/spp_traits_trait_val_lookup.csv'))
write_csv(new_taxa_df, here('update/spp_vuln_framework_traits/spp_traits_taxa_lookup.csv'))
write_csv(new_gapfill_df, here('update/spp_vuln_framework_traits/spp_traits_gapfill_levels.csv'))
```

### Vulnerability score outputs

Separate out lookup tables

```{r}
vuln_junction <- fread(here_anx('3_vuln_score_traits/spp_vuln_from_traits_all_scores.csv'))
vuln_stressor_lookup <- fread(here_anx('3_vuln_score_traits/spp_vuln_from_traits_str.csv'))
vuln_tx_lookup <- fread(here_anx('3_vuln_score_traits/spp_vuln_from_traits_tx.csv')) %>%
  mutate(sciname = species)

new_vuln_df <- vuln_junction %>%
  left_join(vuln_tx_lookup, by = 'vuln_tx_id') %>%
  left_join(vuln_stressor_lookup, by = 'vuln_str_id') %>%
  left_join(update_sciname_aphia_df, by = c('species' = 'sciname')) %>%
  filter(!is.na(aphia_id))

new_vuln_taxa_df <- new_vuln_df %>%
  select(aphia_id, sciname = valid_sciname, taxon) %>%
  distinct() %>% arrange(aphia_id)

new_vuln_str_df <- new_vuln_df %>%
  select(vuln_str_id, stressor) %>%
  distinct()

```

For the grand junction table, when resolving species names to Aphia IDs, some species are now combined that had been separate.  How to reconcile?

* For sensitivity and adaptive capacity, take mean value
* for exposure mod, only 2 species have this issue; take the max (exposed).
* then calc the vulnerability per formula, reusing code from script 6.

```{r}
spp_scores_mean <- new_vuln_df %>%
  group_by(aphia_id, vuln_str_id) %>%
  summarize(sens_score = mean(sens_score), 
            adcap_gen_score = mean(adcap_gen_score), 
            adcap_spec_score = mean(adcap_spec_score), 
            exposure_mod = max(exposure_mod), 
            mean_vuln = mean(vuln),
            .groups = 'drop')

### Identify ref values as max value or 1
ref_values <- spp_scores_mean %>%
  group_by(vuln_str_id) %>%
  summarize(max_sens = max(sens_score, 1), 
              ### the 1 ensures that stressors with very low scores 
              ### don't get normalized by a low reference
            max_adcap_gen = max(adcap_gen_score, 1),
            max_adcap_spec = max(adcap_spec_score, 1), 
            .groups = 'drop')

### Normalize values against reference for each category
rescale_vals <- spp_scores_mean %>%
  left_join(ref_values, by = 'vuln_str_id') %>%
  mutate(sens_rescale = sens_score / max_sens,
         acg_rescale  = adcap_gen_score / max_adcap_gen,
         acs_rescale =  adcap_spec_score / max_adcap_spec)

### Calc vulnerability per formula, then rescale to 0-1 
new_vuln_calc <- rescale_vals %>%
  mutate(vuln_raw = sens_rescale / (1 + acg_rescale + acs_rescale) * exposure_mod,
         vuln = vuln_raw / max(vuln_raw))


# asdf <- janitor::get_dupes(new_vuln_df, aphia_id, vuln_str_id)
# qwer <- asdf %>% group_by(aphia_id, vuln_str_id) %>% filter(n_distinct(exposure_mod) > 1)
# table(qwer$taxon) ### 1 elasmo spp, 1 sponge spp

new_vuln_junction <- spp_scores_mean %>%
  left_join(new_vuln_calc %>% select(aphia_id, vuln_str_id, vuln), by = c('aphia_id', 'vuln_str_id'))

check_diffs <- new_vuln_junction %>%
  filter(vuln != mean_vuln)
### 16 spp with differences in vulnerability
```

Write out new tables

```{r}
dir.create(here('update/spp_vuln_framework_scores'))
write_csv(new_vuln_junction %>% select(-mean_vuln), here('update/spp_vuln_framework_scores/spp_vuln_junction.csv'))
write_csv(new_vuln_str_df, here('update/spp_vuln_framework_scores/spp_vuln_stressor_lookup.csv'))
write_csv(new_vuln_taxa_df, here('update/spp_vuln_framework_scores/spp_vuln_taxa_lookup.csv'))
```

